Conditional resampling improves calibration \\ in single cell CRISPR screen analysis

### 2020-11-29 sceptre_paper/manuscript/manuscript_v3.tex
update manuscript

%% Template for a preprint Letter or Article for submission
%% to the journal Nature.
%% Written by Peter Czoschke, 26 February 2004
%%

\documentclass{nature}

%% make sure you have the nature.cls and naturemag.bst files where
%% LaTeX can find them

\usepackage{graphicx}	
\usepackage{amssymb, amsmath, bbm}
\usepackage{geometry}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}
\newcommand\independent{\perp \!\!\! \perp}
% \usepackage{chngcntr}

@@

%\usepackage[minnames = 3, maxnames=3, url = false, doi = false, isbn = false, sorting = none, backend=bibtex]{biblatex}
%\addbibresource{/home/ekatsevi/Dropbox/Research/Literature/library.bib}

@@

%% Notice placement of commas and superscripts and use of &
%% in the author list

@@

\begin{abstract}
Single cell CRISPR screens are an emerging biotechnology promising unprecedented insights into gene regulation. However, the analysis of these screens presents significant statistical challenges. For example, technical factors like sequencing depth impact not only expression measurement but also perturbation detection, creating a confounding effect. We demonstrate on two recent large-scale single cell CRISPR screens how these challenges cause calibration issues among existing analysis methods. To address these challenges, we propose SCEPTRE: analysis of single cell perturbation screens via conditional resampling. This methodology, designed to avoid calibration issues due to technical confounders and expression model misspecification, infers associations between perturbations and expression by resampling the former according to a working model for perturbation detection probability in each cell. Applying SCEPTRE to the CRISPR screen data yields excellent calibration on negative control perturbations and 200 regulatory relationships not found in the original study, many of which are supported by existing functional data.
\end{abstract}

@@

Single cell CRISPR screens (implemented independently as Perturb-seq\cite{Dixit2016, Adamson2016}, CRISP-seq\cite{Jaitin2016}, and CROP-seq\cite{Datlinger2017}) facilitate unprecedented insights into gene regulation by measuring the impact of precisely targeted perturbations on the whole transcriptome. These pooled screens deliver a library of CRISPR perturbations, targeting either genes or non-coding regulatory elements, to a population of cells and then employ single cell RNA sequencing (scRNA-seq) to identify the perturbations each cell received and measure its transcriptional response. Single cell CRISPR screens have been scaled up through multiplexing to systematically map gene-enhancer relationships\cite{Xie2017,Gasperini2019,Xie2019}, applied using combinatorial perturbation libraries to study genetic interactions\cite{Norman2019}, and extended to single cell protein expression readout\cite{Mimitou2019,Frangieh2020}. Given their scale, resolution, and versatility, these assays are likely to be a major driver of discovery in functional genomics and its applications to the molecular understanding of human disease in the coming decade\cite{Gasperini2020}.

@@

In this work, we explore statistical challenges of single cell CRISPR screens and present a novel analysis methodology to address them. In addition to known scRNA-seq analysis challenges such as expression normalization and discreteness\cite{Townes2019,Hafemeister2019,Svensson2020}, we identify a key challenge that sets CRISPR screens apart from traditional differential expression experiments: the ``treatment"---in this case the presence of a CRISPR perturbation in a given cell---is subject to measurement error\cite{Dixit2016,Hill2018,Replogle2020}. In fact, underlying this measurement error are the same technical factors contributing to errors in the measurement of gene expression, including sequencing depth and batch effects. These technical factors therefore act as confounders, invalidating traditional nonparametric calibration approaches such as those based on ranks or permutations. Indeed, the symmetry among cells assumed by these approaches breaks down in the presence of technical factors varying across cells. On the other hand, parametric modeling of single cell expression data is also fraught with challenges and remains an active area of research.

To address these challenges, we propose SCEPTRE (analysis of Single CEll PerTurbation screens via conditional REsampling; pronounced ``scepter''). This resampling methodology is based on the conditional randomization test\cite{CetL16}, a recently proposed statistical methodology which, like traditional nonparametric methods, does not rely on correct specification of the outcome (expression) model. Unlike traditional nonparametric methods, SCEPTRE employs a resampling scheme that explicitly accounts for heterogeneity in technical factors among cells. We analyze three datasets -- that of Gasperini et al.\cite{Gasperini2019}, that of Xie et al.\cite{Xie2019}, and a synthetic dataset -- to illustrate the aforementioned analysis challenges and the performance of the proposed methodology. SCEPTRE demonstrates excellent calibration and reveals many novel regulatory relationships supported by a variety of existing sources of functional evidence.

@@

Gasperini et al.\ and Xie et al.\ computed $p$-values for gRNA-gene pairs
to test for associations between candidate enhancers and genes. Gasperini et al.\ obtained $p$-values from a DESeq2\cite{Love2014}-inspired negative binomial regression analysis implemented in Monocle2.\cite{Qiu2017} Xie et al., in contrast, obtained $p$-values from Virtual FACS, a nonparametric rank-based method proposed by the authors.\cite{Xie2017} Gasperini et al.\ and Xie et al.\  observed miscalibration in their respective sets of $p$-values. Gasperini et al.\ examined the distribution of negative binomial $p$-values obtained from pairing each protein-coding gene with each of 50 non-targeting control gRNAs. These ``null'' $p$-values exhibited inflation, deviating substantially from the expected uniform distribution (Figure \ref{fig:analysis-challenges}a, top panel). Xie et al.\ examined the distribution of Virtual FACS $p$-values obtained from pairing gRNAs that targeted an enhancer of gene \textit{ARL15}, ARL15-enh, with all protein-coding genes and long noncoding RNAs (lncRNAs) genome-wide. As ARL15-enh is expected to modulate the expression of ARL15 and neighboring lncRNAs only, these $p$-values can be used to assess model calibration. Xie et al.\
 found these $p$-values to be miscalibrated, with those corresponding to protein-coding genes exhibiting conservative bias (Figure \ref{fig:analysis-challenges}a, bottom panel) and those corresponding to lncRNAs exhibiting liberal bias (not depicted).

Confounding likely is a cause of the miscalibration noted by Gasperini et al.\ and Xie et al. Despite the use of targeted amplification protocols for gRNA detection\cite{Hill2018}, not all transduced gRNAs are detected in single cell CRISPR screen experiments. Indeed, the total number of gRNAs detected in a cell increases with the sequencing depth, defined as the total number of mRNA UMIs detected per cell ($\rho = 0.35, p < 10^{-15}$ in Gasperini et al.\ data; $\rho = 0.25, p < 10^{-15}$ in Xie et al.\ data; Figures \ref{fig:analysis-challenges}b-c). Therefore, sequencing depth and other technical factors induce correlations between the measured gRNA presence and gene expression, even in the absence of a regulatory relationship (Figure \ref{fig:analysis-challenges}d). This confounding effect, if not explicitly accounted for, results in test miscalibration. Confounding is especially problematic for nonparametric methods like Virtual FACS, which implicitly (and incorrectly) assume that there are no technical factors that impact both gRNA detection and gene expression measurement.

@@

Gasperini et al.\ and Xie et al.\ adjusted their analyses to remedy the observed calibration problems. Gasperini et al.\ calibrated the candidate enhancer $p$-values against the distribution of all gene-negative control $p$-values instead of the uniform distribution. The resulting ``empirical $p$-values'' were used for determining significance. Along similar lines, Xie et al.\ calibrated the Virtual FACS $p$-values against a simulated set of gene-specific ``null'' $p$-values. These analysis adjustments, unfortunately, did not satisfactorily improve calibration. The adjustment of Gasperini et al., for example, lead to overcorrection for some gene-enhancer pairs (false negatives) and undercorrection for others (false positives) (Figure 1s). Methods that avoid reliance on ad hoc calibration fixes might produce more consistently calibrated $p$-values across entire CRISPR screen datasets.

\begin{figure}[h!]
\includegraphics[width = \textwidth]{figures/Figure1/Figure1_cropped.pdf}
\caption{\textbf{CRISPR screen analysis challenges can lead to false positives and false negatives}. \textbf{a}, (Top) QQ-plot of Gasperini et al.\ p-values for all pairs of genes and negative control gRNAs. (Bottom) QQ-plot of Xie et al.\ p-values for all pairs of genes and ARL15-enh-targeting gRNAs (\textit{ARL15} gene omitted). These sets of $p$-values deviate from the expected uniform distribution, indicating test miscalibration. \textbf{b-d}, Sequencing depth (total mRNA UMIs per cell) impacts gRNA detection and observed gene expression levels in both Gasperini et al.\ (\textbf{b}) and Xie et al.\ (\textbf{c}) data. Thus, sequencing depth acts as a confounder (\textbf{d}). This analysis challenge distinguishes CRISPR screens from traditional differential expression applications. \textbf{e}, Monocle2 estimates the dispersion of each gene by projecting each gene's raw dispersion estimate onto the fitted raw dispersion-mean expression curve. This estimation procedure leads to miscalibration for high-dispersion genes.  % \textbf{b-c}, Sequencing depth (total mRNA UMIs per cell) impacts gRNA detection (b) and observed expression levels, and therefore acts as a confounder (c). This analysis challenge distinguishes CRISPR screens from traditional differential expression applications. }
}
 \label{fig:analysis-challenges}
\end{figure}
\clearpage

@@

We reanalyzed the Gasperini et al.\ data using the improved negative binomial regression approach. In addition to sequencing depth, we included as covariates in the regression model the total number of expressed genes per cell, as well as the technical factors that Gasperini et al.\ accounted for in their analysis (namely, total number of gRNAs detected per cell, percentage of transcripts mapped to mitochondrial genes, and sequencing batch). We also applied the improved negative binomial approach to the Xie et al.\ ARL15-enh data. In the latter analysis we included sequencing depth, number of detected gRNA UMIs per cell, and sequencing batch as covariates. We found that the improved negative binomial model was better calibrated than the Monocle2 negative binomial model on the Gasperini data (Figure 3c; improved negative binomial denoted ``Hafemeister NB''). However, on both the Gasperini et al.\ data and Xie et al.\ data, the improved negative binomial model showed clear signs of p-value inflation (Figure 3c-d).

@@

% The difficulty of applying parametric models to single cell data highlights the appeal of nonparametric approaches for this purpose.
\clearpage
\paragraph{SCEPTRE: Analysis of single cell perturbation screens via conditional resampling.}

@@

\clearpage
\thispagestyle{empty}
\begin{figure}[h!]
	\includegraphics[width = \textwidth]{figures/Figure2/Figure2.png}
	\caption{\textbf{SCEPTRE: Analysis of single cell perturbation screens via conditional resampling.} A schematic and outline of the SCEPTRE methodology for one gene and one gRNA is shown, entailing the application of the conditional randomization test\cite{CetL16} to single cell CRISPR screens. The idea is to assign to every cell the probability of gRNA observation based on its technical factors, and then to build a null distribution for the negative binomial $z$-value by independently resampling the gRNA indicator for each cell according to these probabilities to form ``negative control" data sets. A skew-$t$ distribution is fit to the resulting histogram to obtain precise $p$-values based on a limited number of resamples, against which the original NB $z$-value is compared. The dashed line shows the standard normal distribution, against which the NB $z$-value would normally be compared.}
	\label{fig:method}
\end{figure}
\clearpage

@@

We assessed the calibration, sensitivity, and specificity of SCEPTRE in several settings in which the ground truth was known. First, we demonstrated the calibration of SCEPTRE in a small, proof-of-concept simulation study (Figure \ref{fig:ground_truth}a). We considered a class of negative binomial regression models with dispersion $\alpha$ and two technical covariates (sequencing depth and batch). We simulated expression data for a single gene in 1000 cells using four models selected from this class: the first with $\alpha = 1$, the second with $\alpha = 0.25$, the third with $\alpha = 5$, and the last with $\alpha = 1,$ but with 25$\%$ zero-inflation. We also simulated negative control gRNA data using a logistic regression model with the same covariates as in the gene expression model. We assessed the calibration of three methods across the four simulated datasets: SCEPTRE, improved negative binomial regression, and scMAGeCK-LR,\cite{Yang2020} a recently-proposed, permutation-based nonparametric method. To assess the impact of model misspecification on SCEPTRE and the improved negative binomial method (on which SCEPTRE relies), we fixed the dispersion of the negative binomial method to 1 across all four simulated datasets. We found that the negative binomial method works as expected when the model is correctly specified but breaks down in all three cases of model misspecification. scMAGeCK-LR exhibits poor calibration across all simulated datasets, likely because, as a traditional nonparametric method, it fails to adequately account for confounders. Finally, SCEPTRE is well-calibrated regardless of expression model misspecification and confounder presence.

@@

Last, we assessed the sensitivity of SCEPTRE using the Gasperini et al.\ positive control data and confirmatory bulk RNA-seq data produced by Xie et al. We applied SCEPTRE to the 381 positive control gRNA-gene pairs assayed in Gasperini et al.\ (Figure~\ref{fig:ground_truth}d), where each gRNA targeted the transcription start site of the corresponding gene. The SCEPTRE $p$-values for these positive controls are highly significant, and in particular, more significant than the original empirical $p$-values, indicating greater sensitivity. We note that the empirical correction employed by Gasperini et al.\ limits the accuracy of $p$-values to about $10^{-6}$, explaining at least part of the difference. Finally, we compared the SCEPTRE results for ARL15-enh to the results of an arrayed CRISPR screen of ARL15-enh in which populations of perturbed cells were compared to populations of unperturbed cells via bulk RNA-seq.\cite{Xie2019} Both SCEPTRE and the bulk RNA-seq differential expression analysis rejected gene \textit{ARL15} (and only gene \textit{ARL15}) at an FDR of 0.1 after a Benjamini-Hochberg correction, increasing our confidence in sensitivity (and specificity) of SCEPTRE (Figure \ref{fig:ground_truth}e).

\newpage
\begin{figure}[h!]
	\includegraphics[width = \textwidth]{figures/Figure3/Figure3_cropped.pdf}
	\caption{\textbf{SCEPTRE demonstrates excellent calibration, specificity, and sensitivity when the ground truth is known.}}
	\label{fig:ground_truth}
 \end{figure}
\clearpage

% \clearpage
% \thispagestyle{empty} 
% \addtocounter{page}{-1}
%\begin{figure}[h!]
%	\includegraphics[width = \textwidth]{figures/Figure3/Figure3.png}
%	\caption{\textbf{SCEPTRE improves on the sensitivity and specificity of the parametric approach by correcting for expression model misspecification.} \textbf{a,} Numerical simulation comparing three ways of calibrating a negative binomial test statistic (parametric, permutation, conditional resampling), when the test statistic is based on the correct model or contains expression model misspecifications. Only the conditional resampling approach maintains calibration despite model misspecification. \textbf{b,c,} Application of SCEPTRE to Gasperini et al negative control gRNAs, comparing all gene-NTC pairs to the uniform distribution (b) or breaking down by NTC (c); details as in Figure~\ref{fig:calibration-comparison-all}a,c. The matching of SCEPTRE's colored region with the gray one in panel c, together with the overall near-uniformity in panel b, demonstrates its excellent calibration on real negative control data. \textbf{d,} SCEPTRE $p$-values for TSS-targeting controls generally more significant than the original empirical $p$-values; dashed line is the $45^\circ$ line. Note: $p$-values truncated for visualization, and empirical $p$-values in panel d at least 1/538,560 $\approx$ 2$\times 10^{-6}$ by construction.}
%	\label{fig:calibration}
% \end{figure}
\clearpage

@@

We applied SCEPTRE to the 84,595 gene-enhancer pairs considered in Gasperini et al., encompassing 10,560 genes and 5,779 candidate enhancers. We applied the Benjamini-Hochberg correction at level 0.1 (the same level used in the original analysis) to the $p$-values obtained for all of these candidate pairs, obtaining a total of 563 gene-enhancer pairs. By comparison, Gasperini et al. found 470 high-confidence pairs. Comparing the SCEPTRE $p$-values against the original empirical $p$-values (Figure~\ref{fig:application}a), we see that the two often diverge substantially. Our analysis found 200 gene-enhancer pairs that the original analysis did not, while 107 were found only by the original analysis. Many of the discoveries found only in the original analysis show signs of the $p$-value inflation observed in Figure~\ref{fig:analysis-challenges}.

@@

We also found that the total set of gene-enhancer pairs discovered was better enriched for regulatory biological signals, including HI-C and ChIP-seq. Gene-enhancer pairs discovered by SCEPTRE tended to be physically closer to each other on average that those discovered in the original analysis (Figure~\ref{fig:application}c). Furthermore, 74.4\% of SCEPTRE's 563 gene-enhancer pairs fell in the same topologically associating domain (TAD), compared to 71.1\% of the original 470 pairs. We also repeated Gasperini et al.'s contact frequency enrichment analysis for those pairs falling in the same TAD (Figure~\ref{fig:application}d). We found similar levels of contact frequency enrichment, despite the fact that SCEPTRE discovered 85 more gene-enhancer pairs in the same TADs. Finally, we repeated the ChIP-seq enrichment analysis of Gasperini et al. for seven cell-type relevant transcription factors and H3K27ac. We quantified enrichment of ChIP-seq signal in paired enhancers as the odds ratio that a candidate enhancer is paired to a gene, comparing those falling in the top quintile of candidate enhancers by ChIP-seq signal and those not overlapping a ChIP-seq peak at all. We find improved enrichment for each of the eight ChIP-seq targets (Figures~\ref{fig:application}e and \ref{fig:chipseq-enrichment-details}).

>>>>

\documentclass{nature}
\usepackage{graphicx}	
\usepackage{amssymb, amsmath, bbm}
\usepackage{geometry}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}
\newcommand\independent{\perp \!\!\! \perp}
\usepackage{capt-of}
% \usepackage{chngcntr}

@@

@@

@@

\begin{abstract}
Single cell CRISPR screens are an emerging biotechnology promising unprecedented insights into gene regulation. However, the analysis of these screens presents significant statistical challenges. For example, technical factors like sequencing depth impact not only expression measurement but also perturbation detection, creating a confounding effect. We demonstrate on two recent large-scale single cell CRISPR screens how these challenges cause calibration issues among existing analysis methods. To address these challenges, we propose SCEPTRE: analysis of single cell perturbation screens via conditional resampling. This methodology, designed to avoid calibration issues due to technical confounders and expression model misspecification, infers associations between perturbations and expression by resampling the former according to a working model for perturbation detection probability in each cell. SCETPRE demonstrates excellent calibration, sensitivity, and specificity on the CRISPR screen data and yields 200 new regulatory relationships, many of which are supported by existing functional data.
\end{abstract}

@@

Single cell CRISPR screens (implemented independently as Perturb-seq\cite{Dixit2016, Adamson2016}, CRISP-seq\cite{Jaitin2016}, and CROP-seq\cite{Datlinger2017}) facilitate unprecedented insights into gene regulation by measuring the impact of precisely targeted perturbations on the whole transcriptome. These pooled screens deliver a library of CRISPR perturbations, targeting either genes or non-coding regulatory elements, to a population of cells and then employ single cell RNA sequencing (scRNA-seq) to identify the perturbations each cell received and measure its transcriptional response. Single cell CRISPR screens have been scaled up through multiplexing to systematically map gene-enhancer relationships,\cite{Xie2017,Gasperini2019,Xie2019} applied using combinatorial perturbation libraries to study genetic interactions,\cite{Norman2019} and extended to single cell protein expression readout.\cite{Mimitou2019,Frangieh2020} Given their scale, resolution, and versatility, these assays are likely to be a major driver of discovery in functional genomics and its applications to the molecular understanding of human disease in the coming decade.\cite{Gasperini2020}

@@

In this work, we explore statistical challenges of single cell CRISPR screens and present a novel analysis methodology to address them. In addition to known scRNA-seq analysis challenges such as expression normalization and discreteness,\cite{Townes2019,Hafemeister2019,Svensson2020} we identify a key challenge that sets CRISPR screens apart from traditional differential expression experiments: the ``treatment"---in this case the presence of a CRISPR perturbation in a given cell---is subject to measurement error.\cite{Dixit2016,Hill2018,Replogle2020} In fact, underlying this measurement error are the same technical factors contributing to errors in the measurement of gene expression, including sequencing depth and batch effects. These technical factors therefore act as confounders, invalidating traditional nonparametric calibration approaches such as those based on ranks or permutations. Indeed, the symmetry among cells assumed by these approaches breaks down in the presence of technical factors varying across cells. On the other hand, parametric modeling of single cell expression data is also fraught with challenges and remains an active area of research.

To address these challenges, we propose SCEPTRE (analysis of Single CEll PerTurbation screens via conditional REsampling; pronounced ``scepter''). This resampling methodology is based on the conditional randomization test,\cite{CetL16} a recently proposed statistical methodology which, like traditional nonparametric methods, does not rely on correct specification of the outcome (expression) model. Unlike traditional nonparametric methods, SCEPTRE employs a resampling scheme that explicitly accounts for heterogeneity in technical factors among cells. We analyze data from two recent single cell pooled CRISPR screen experiments--one produced by Gasperini et al.\cite{Gasperini2019} and one produced by Xie et al.\cite{Xie2019}--to illustrate the aforementioned analysis challenges and the performance of the proposed methodology. SCEPTRE demonstrates excellent calibration, specificity, and sensitivity on the data and reveals many novel regulatory relationships supported by a variety of existing sources of functional evidence.

@@

Gasperini et al.\ and Xie et al.\ computed $p$-values for gRNA-gene pairs
to test for associations between candidate enhancers and genes. Gasperini et al.\ obtained $p$-values from a DESeq2\cite{Love2014}-inspired negative binomial regression analysis implemented in Monocle2.\cite{Qiu2017} Xie et al., in contrast, obtained $p$-values from Virtual FACS, a nonparametric rank-based method proposed by the authors.\cite{Xie2017} Gasperini et al.\ and Xie et al.\  observed miscalibration in their respective sets of $p$-values. Gasperini et al.\ examined the distribution of negative binomial $p$-values obtained from pairing each protein-coding gene with each of 50 non-targeting control (or negative control) gRNAs. These ``null'' $p$-values exhibited inflation, deviating substantially from the expected uniform distribution (Figure \ref{fig:analysis-challenges}a, top panel). Xie et al.\ examined the distribution of Virtual FACS $p$-values obtained from pairing gRNAs that targeted an enhancer of gene \textit{ARL15}, ARL15-enh, with all protein-coding genes and long noncoding RNAs (lncRNAs) genome-wide. As ARL15-enh is expected to modulate the expression of ARL15 and neighboring lncRNAs only, these $p$-values can be used to assess model calibration. Xie et al.\
 found these $p$-values to be miscalibrated, with those corresponding to protein-coding genes exhibiting conservative bias (Figure \ref{fig:analysis-challenges}a, bottom panel) and those corresponding to lncRNAs exhibiting liberal bias (not depicted).

Confounding likely is a cause of the miscalibration noted by Gasperini et al.\ and Xie et al. Despite the use of targeted amplification protocols for gRNA detection\cite{Hill2018}, not all transduced gRNAs are detected in single cell CRISPR screen experiments. Indeed, the total number of gRNAs detected in a cell increases with the sequencing depth, defined as the total number of mRNA UMIs detected per cell ($\rho = 0.35, p < 10^{-15}$ in Gasperini et al.\ data; $\rho = 0.25, p < 10^{-15}$ in Xie et al.\ data; Figures \ref{fig:analysis-challenges}b-c). Therefore, sequencing depth and other technical factors induce correlations between the measured gRNA presence and gene expression, even in the absence of a regulatory relationship (Figure \ref{fig:analysis-challenges}d). This confounding effect, if not explicitly accounted for, results in test miscalibration. Confounding is especially problematic for traditional nonparametric methods like Virtual FACS, which implicitly (and incorrectly) assume that there are no technical factors that impact both gRNA detection and gene expression measurement.

@@

Gasperini et al.\ and Xie et al.\ introduced ad hoc adjustments to their analyses to remedy the observed calibration issues. Gasperini et al.\ calibrated the candidate enhancer $p$-values against the distribution of all negative control gRNA $p$-values instead of the uniform distribution. The resulting ``empirical $p$-values'' were used for determining significance. Along similar lines, Xie et al.\ calibrated the Virtual FACS $p$-values against a simulated set of gene-specific ``null'' $p$-values. These analysis adjustments, unfortunately, did not satisfactorily improve calibration. The adjustment employed by Gasperini et al., for example, leads to overcorrection for some gene-enhancer pairs (false negatives) and undercorrection for others (false positives) (Figure \ref{fig:gasp_correction}). Methods that avoid reliance on ad hoc calibration fixes likely would produce more consistently calibrated $p$-values across entire CRISPR screen datasets.

\begin{figure}[h!]
\includegraphics[width = \textwidth]{figures/Figure1/Figure1_cropped.pdf}
\caption{\textbf{CRISPR screen analysis challenges can lead to false positives and false negatives}. \textbf{a}, (Top) QQ-plot of Gasperini et al.\ p-values for all pairs of genes and negative control gRNAs. (Bottom) QQ-plot of Xie et al.\ p-values for all pairs of genes and ARL15-enh-targeting gRNAs (\textit{ARL15} gene omitted). These sets of $p$-values deviate from the expected uniform distribution, indicating test miscalibration. \textbf{b-d}, Sequencing depth (total mRNA UMIs per cell) impacts gRNA detection and observed gene expression levels in both Gasperini et al.\ (b) and Xie et al.\ (c) data. Thus, sequencing depth acts as a confounder (d). This analysis challenge distinguishes CRISPR screens from traditional differential expression applications. \textbf{e}, Monocle2 estimates the dispersion of each gene by projecting each gene's raw dispersion estimate onto the fitted raw dispersion-mean expression curve. This estimation procedure leads to miscalibration for high-dispersion genes.
}
 \label{fig:analysis-challenges}
\end{figure}
\clearpage

@@

We reanalyzed the Gasperini et al.\ data using the improved negative binomial regression approach. In addition to sequencing depth, we included as covariates in the regression model the total number of expressed genes per cell, as well as the technical factors that Gasperini et al.\ accounted for in their analysis (namely, total number of gRNAs detected per cell, percentage of transcripts mapped to mitochondrial genes, and sequencing batch). We also applied the improved negative binomial approach to the Xie et al.\ ARL15-enh data. We included sequencing depth, number of detected gRNA UMIs per cell, and sequencing batch as covariates in the analysis of the latter dataset. The improved negative binomial method exhibited better calibration than the Monocle negative binomial method on the Gasperini data (Figure \ref{fig:ground_truth}b). However, on both the Gasperini et al.\ data and Xie et al.\ data, the improved negative binomial method showed clear signs of p-value inflation (Figure \ref{fig:ground_truth}b-c).

@@

\paragraph{SCEPTRE: Analysis of single cell perturbation screens via conditional resampling.}

@@

\thispagestyle{empty}
\begin{figure}[h!]
	\includegraphics[width = \textwidth]{figures/Figure2/Figure2.png}
	\caption{\textbf{SCEPTRE: Analysis of single cell perturbation screens via conditional resampling.} A schematic and outline of the SCEPTRE methodology for one gene and one gRNA is shown, entailing the application of the conditional randomization test\cite{CetL16} to single cell CRISPR screens. The idea is to assign to every cell the probability of gRNA observation based on its technical factors, and then to build a null distribution for the negative binomial $z$-value by independently resampling the gRNA indicator for each cell according to these probabilities to form ``negative control" data sets. A skew-$t$ distribution is fit to the resulting histogram to obtain precise $p$-values based on a limited number of resamples, against which the original NB $z$-value is compared. The dashed line shows the standard normal distribution, against which the NB $z$-value would normally be compared.}
	\label{fig:method}
\end{figure}

@@

We assessed the calibration, sensitivity, and specificity of SCEPTRE in several settings in which the ground truth was known. First, we demonstrated the calibration of SCEPTRE in a small, proof-of-concept simulation study (Figure \ref{fig:ground_truth}a). We considered a class of negative binomial regression models with dispersion $\alpha$ and two technical covariates (sequencing depth and batch). We simulated expression data for a single gene in 1000 cells using four models selected from this class: the first with $\alpha = 1$, the second with $\alpha = 0.25$, the third with $\alpha = 5$, and the last with $\alpha = 1,$ but with 25$\%$ zero-inflation. We also simulated negative control gRNA data using a logistic regression model with the same covariates as the gene expression model. We assessed the calibration of three methods across the four simulated datasets: SCEPTRE, improved negative binomial regression, and scMAGeCK-LR,\cite{Yang2020} a recently-proposed, permutation-based nonparametric method. To assess the impact of model misspecification on SCEPTRE and the improved negative binomial method (on which SCEPTRE relies), we fixed the dispersion of the negative binomial method to 1 across all four simulated datasets. We found that the negative binomial method works as expected when the model is correctly specified but breaks down in all three cases of model misspecification. scMAGeCK-LR exhibits poor calibration across all simulated datasets, likely because, as a traditional nonparametric method, it fails to adequately account for confounders. Finally, SCEPTRE is well-calibrated regardless of expression model misspecification and confounder presence.

@@

Last, we assessed the sensitivity of SCEPTRE using the Gasperini et al.\ positive control data and confirmatory bulk RNA-seq data produced by Xie et al. We applied SCEPTRE to the 381 positive control gRNA-gene pairs assayed in Gasperini et al.\ (Figure~\ref{fig:ground_truth}d), where each gRNA targeted the transcription start site of the corresponding gene. The SCEPTRE $p$-values for these positive controls are highly significant, and in particular, more significant than the original empirical $p$-values, indicating greater sensitivity. We note that the empirical correction employed by Gasperini et al.\ limits the accuracy of $p$-values to about $10^{-6}$, explaining at least part of the difference. Finally, we compared the SCEPTRE ARL15-enh results to the results of an arrayed CRISPR screen of ARL15-enh in which populations of perturbed cells were compared to populations of unperturbed cells via bulk RNA-seq.\cite{Xie2019} Both SCEPTRE and the bulk RNA-seq differential expression analysis rejected gene \textit{ARL15} (and only gene \textit{ARL15}) at an FDR of 0.1 after a Benjamini-Hochberg correction, increasing our confidence in sensitivity (and specificity) of SCEPTRE (Figure \ref{fig:ground_truth}e).

\clearpage
\begin{figure*}[h!]
	\includegraphics[width = \textwidth]{figures/Figure3/Figure3_cropped.pdf}
 \end{figure*}
\captionof{figure}[short caption here]{
	\label{fig:ground_truth}
	\textbf{SCEPTRE demonstrates excellent calibration, specificity, and sensitivity under known ground truth.} \textbf{a,} Numerical simulation comparing three methods -- SCEPTRE, improved negative binomial regression, and scMAGeCK-LR -- on four simulated datasets. Only SCEPTRE maintains calibration despite model misspecification and confounder presence. \textbf{b-c},  	Application of SCEPTRE, improved negative binomial regression, Monocle regression, and Virtual FACS to (b) all negative control gRNAs and all genes in the Gasperini et al.\ data, and (c) all ARL15-enh-targeting gRNAs and all genes in the Xie et al.\ data. Compared to the other methods, SCEPTRE shows excellent calibration. Gene \textit{ARL15} annotated in (c). \textbf{d,} SCEPTRE $p$-values for Gasperini et al.\ TSS-targeting controls are highly significant, and in general, more significant than the original empirical $p$-values. \textbf{e,} Comparison of $p$-values produced by SCEPTRE for ARL15-enh to $p$-values produced by an arrayed, bulk RNA-seq CRISPR screen of ARL15-enh. The results of the two analyses coincide exactly, with both aanlyses rejecting gene \textit{ARL15} (and only gene \textit{ARL15}) after a Benjamini-Hochberg correction. Dotted blue lines, rejection thresholds.}
\noindent\hrulefill

@@

We applied SCEPTRE to the 84,595 gene-enhancer pairs considered in Gasperini et al., encompassing 10,560 genes and 5,779 candidate enhancers. We applied the Benjamini-Hochberg correction at level 0.1 (the same level used in the original analysis) to the $p$-values obtained for all of these candidate pairs, obtaining a total of 563 gene-enhancer pairs. By comparison, Gasperini et al. found 470 high-confidence pairs. Comparing the SCEPTRE $p$-values against the original empirical $p$-values (Figure~\ref{fig:application}a), we see that the two often diverge substantially. SCEPTRE found 200 gene-enhancer pairs that the original analysis did not, while 107 were found only by the original analysis. Many of the discoveries found only in the original analysis show signs of the $p$-value inflation observed in Figure~\ref{fig:gasp_correction}.

@@

\clearpage
\begin{figure}[h!]
	\includegraphics[width = \textwidth]{figures/Figure4/Figure4_cropped.pdf}
\end{figure}
\captionof{figure}[short caption here]{
	\label{fig:application}
	\textbf{Application of SCEPTRE to Gasperini et al.\ data yields biologically plausible gene-enhancer links.} \textbf{a,} Comparison of the original empirical $p$-values to those obtained from SCEPTRE. The two analysis methods differ substantially, with 200 gene-enhancer links discovered only by SCEPTRE and 107 discovered only by the original. \textbf{b}, Five gene-enhancer pairs discovered by SCEPTRE but not the original analysis, each supported by a whole blood GTEx eQTL or FANTOM enhancer RNA correlation $p$-value. \textbf{c,} Distribution of distances from TSS to upstream paired enhancers. Compared to the original analysis, SCEPTRE pairs genes with nearer enhancers on average. \textbf{d,} For those gene-enhancer pairs falling in the same TAD, the cumulative distribution of the fractional rank of the HI-C interaction frequency compared to other distance-matched loci pairs within the same TAD. SCEPTRE shows similar enrichment despite finding 25\% more within-TAD pairs. Inset table shows gene-enhancer pairs falling in the same TAD. SCEPTRE finds 93 more total pairs, and a higher percentage of pairs fall in the same TAD. \textbf{e,} Enrichment of ChIP-seq signal from seven cell-type relevant transcription factors and one histone mark among paired enhancers. SCEPTRE exhibits greater enrichment across all ChIP-seq targets.}
\noindent\hrulefill

We also found that the total set of gene-enhancer pairs discovered was better enriched for regulatory biological signals, including HI-C and ChIP-seq. Gene-enhancer pairs discovered by SCEPTRE tended to be physically closer to each other on average that those discovered in the original analysis (Figure~\ref{fig:application}c). Furthermore, 74.4\% of SCEPTRE's 563 gene-enhancer pairs fell in the same topologically associating domain (TAD), compared to 71.1\% of the original 470 pairs. We also repeated Gasperini et al.'s contact frequency enrichment analysis for those pairs falling in the same TAD (Figure~\ref{fig:application}d). We found similar levels of contact frequency enrichment, despite the fact that SCEPTRE discovered 85 more gene-enhancer pairs in the same TADs. Finally, we repeated the ChIP-seq enrichment analysis of Gasperini et al.\ for seven cell-type relevant transcription factors and H3K27ac. We quantified enrichment of ChIP-seq signal in paired enhancers as the odds ratio that a candidate enhancer is paired to a gene, comparing those falling in the top quintile of candidate enhancers by ChIP-seq signal and those not overlapping a ChIP-seq peak at all. We find improved enrichment for each of the eight ChIP-seq targets (Figures~\ref{fig:application}e and \ref{fig:chipseq-enrichment-details}).
